---
title: "USA Storm Data Analysis"
output: html_document
---

# USA Storm impacts on health and economics

Load and reduce data to most relevant variables

Data clean up
- Reduce the variables to health and economcis data
- include data 1996-2010 for which complete anual data, 


Date selection criteria
    - Post 1996 the event types utlised by NOO are consistent http://www.ncdc.noaa.gov/stormevents/details.jsp
    


```{r load_data}
stormAll <- read.csv("repdata-data-StormData.csv")

```

### Data Reduction
    - Reduce the variables to health and economcis data
    - include data 1996-2010 for which complete anual data, 


Date selection criteria
    - Post 1996 the event types utlised by NOO are consistent http://www.ncdc.noaa.gov/stormevents/details.jsp
    
```{r reduce_data}
cols <- c('STATE', 'EVTYPE', 'BGN_DATE', 'END_DATE', 'COUNTYNAME','FATALITIES','INJURIES','PROPDMG', 'PROPDMGEXP','CROPDMG','CROPDMGEXP','REMARKS')
stormBig <- subset(stormAll, select = cols)
stormBig$BGN_DATE <- as.POSIXlt(as.character(stormBig$BGN_DATE),format="%m/%d/%Y %H:%M:%S")
stormBig$END_DATE <- as.POSIXlt(as.character(stormBig$END_DATE),format="%m/%d/%Y %H:%M:%S")

startdate <- as.POSIXlt("01/01/1996", format="%m/%d/%Y")
enddate <- as.POSIXlt("12/31/2010", format="%m/%d/%Y")

storm <- subset(stormBig,trunc(stormBig$BGN_DATE,"days") >= startdate)
storm <- subset(storm,trunc(storm$END_DATE,"days") <= enddate)

```
```{r event_types}
storm$EVTYPE <- toupper(storm$EVTYPE)
storm$EVTYPE[grep('TSTM', storm$EVTYPE)] <- 'THUNDERSTORM WIND'
storm$EVTYPE[grep('RIP CURRENT', storm$EVTYPE)] <- 'RIP CURRENT'
storm$EVTYPE[grep('WILD.*FIRE',storm$EVTYPE)] <- 'WILDFIRE'
storm$EVTYPE[grep('HURRICANE',storm$EVTYPE)] <- 'HURRICANE (TYPHOON)'
storm$EVTYPE[grep('TYPHOON',storm$EVTYPE)] <- 'HURRICANE (TYPHOON) '


```

## Health Effects
Reduce data set

```{r health_data}
cols <- c('EVTYPE','FATALITIES','INJURIES')

health <- subset(storm,storm$FATALITIES > 0 | storm$INJURIES > 0 )
health <- subset(health, select = cols)

# Event type consolidation
health$EVTYPE <- toupper(health$EVTYPE)


healthSum <- aggregate(. ~ EVTYPE,data = health, FUN = sum)

orderHealthSummary <- healthSum[order(-healthSum$FATALITIES),]
top10 <- orderHealthSummary[1:10,]
```

```{r plot}
require(reshape)
require(ggplot2)
t10 <- subset(top10,select=cols)
t10<-melt(t10,id=c('EVTYPE'),variable_name = 'HEALTHIMPACT')

p <- ggplot(t10, aes(x=as.factor(t10$EVTYPE), y= value, fill=t10$HEALTHIMPACT)) +
    geom_bar( stat = "identity") +
    coord_flip() 
p <- p + xlab("Event Type")
p <- p +  ylab("Incidents")
p <- p + ggtitle("Top 10 Fatal USA Storm Event Type 1996-2010")
p <- p + guides(fill=guide_legend(title=""))
p


```

```{r old}

cols <- c('STATE', 'EVTYPE', 'BGN_DATE', 'END_DATE', 'COUNTYNAME','FATALITIES','INJURIES','PROPDMG', 'PROPDMGEXP','CROPDMG','CROPDMGEXP','REMARKS')
```