---
title: "USA Storm Data Analysis"
output: html_document
---

# USA Storm impacts on health and economics

## Synopsis

Across the United States, Storm and other severe events have adverse health and economic consequences.

The most 2 most determiental event to population health are Excessive Heat and Tornados between 1996 and 2010. Excessive Heat resulted in almost 1,800 premature deaths, and more than 6000 individuals to require medical treatment. As for injuries counts, Tornados have the greatest impact, 14500 injuries were attributed to Tornados over the 10 year period, and over 900 fatalities.

Across the United States, which types of events have the greatest economic consequences?
## Data Processing


Load and reduce data to most relevant variables

Data clean up
- Reduce the variables to health and economcis data
- include data 1996-2010 for which complete anual data, 


Date selection criteria
    - Post 1996 the event types utlised by NOO are consistent http://www.ncdc.noaa.gov/stormevents/details.jsp
    


```{r load_data}
# load data
stormAll <- read.csv("repdata-data-StormData.csv")

```

### Data Reduction
    - Reduce the variables to health and economcis data
    - include data 1996-2010 for which complete anual data, 


Date selection criteria
    - Post 1996 the event types utlised by NOO are consistent http://www.ncdc.noaa.gov/stormevents/details.jsp
    
```{r reduce_data}
cols <- c('STATE', 'EVTYPE', 'BGN_DATE', 'END_DATE', 'FATALITIES','INJURIES','PROPDMG', 'PROPDMGEXP','CROPDMG','CROPDMGEXP')
stormBig <- subset(stormAll, select = cols)
stormBig$BGN_DATE <- as.POSIXlt(as.character(stormBig$BGN_DATE),format="%m/%d/%Y %H:%M:%S")
stormBig$END_DATE <- as.POSIXlt(as.character(stormBig$END_DATE),format="%m/%d/%Y %H:%M:%S")

startdate <- as.POSIXlt("01/01/1996", format="%m/%d/%Y")
enddate <- as.POSIXlt("12/31/2010", format="%m/%d/%Y")

storm <- subset(stormBig,trunc(stormBig$BGN_DATE,"days") >= startdate)
storm <- subset(storm,trunc(storm$END_DATE,"days") <= enddate)

```

### Event Type Clean up
```{r event_types}
storm$EVTYPE <- toupper(storm$EVTYPE)
storm$EVTYPE[grep('TSTM', storm$EVTYPE)] <- 'THUNDERSTORM WIND'
storm$EVTYPE[grep('RIP CURRENT', storm$EVTYPE)] <- 'RIP CURRENT'
storm$EVTYPE[grep('WILD.*FIRE',storm$EVTYPE)] <- 'WILDFIRE'
storm$EVTYPE[grep('HURRICANE',storm$EVTYPE)] <- 'HURRICANE (TYPHOON)'
storm$EVTYPE[grep('TYPHOON',storm$EVTYPE)] <- 'HURRICANE (TYPHOON) '

```

### Exponent Clean up
```{r exponent_cleanup}
storm$CROPDMGEXP <- toupper(storm$CROPDMGEXP)
storm$PROPDMGEXP <- toupper(storm$PROPDMGEXP)

storm$CROPMULTIPLER[storm$CROPDMGEXP == ""] <- 1
storm$CROPMULTIPLER[storm$CROPDMGEXP == "H"] <- 100
storm$CROPMULTIPLER[storm$CROPDMGEXP == "K"] <- 1000
storm$CROPMULTIPLER[storm$CROPDMGEXP == "M"] <- 1000000
storm$CROPMULTIPLER[storm$CROPDMGEXP == "B"] <- 1000000000

storm$PROPMULTIPLER[storm$PROPDMGEXP == ""] <- 1
storm$PROPMULTIPLER[storm$PROPDMGEXP == "H"] <- 100
storm$PROPMULTIPLER[storm$PROPDMGEXP == "K"] <- 1000
storm$PROPMULTIPLER[storm$PROPDMGEXP == "M"] <- 1000000
storm$PROPMULTIPLER[storm$PROPDMGEXP == "B"] <- 1000000000

storm$CROPDMG[is.na(storm$CROPDMG)] <- 0
storm$PROPDMG[is.na(storm$PROPDMG)] <- 0

storm$CROP <- storm$CROPDMG*storm$CROPMULTIPLER
storm$PROP <- storm$PROPDMG*storm$PROPMULTIPIER


```
## Results

### Health Effects

The health impacts of severe events can be quanitified according to fatalities and 
injuries. Fatalities are more concerning that injuries, accordingly, only fatalities are considered
for ordering of the events with the greatest impact. While both fatalities and injuries have been presented in visual summary
```{r health_data}
cols <- c('EVTYPE','FATALITIES','INJURIES')

health <- subset(storm,storm$FATALITIES > 0 | storm$INJURIES > 0 )
health <- subset(health, select = cols)

healthSum <- aggregate(. ~ EVTYPE,data = health, FUN = sum)
# Order by Fatal
orderHealthSummary <- healthSum[order(-healthSum$FATALITIES),]
top10 <- orderHealthSummary[1:10,]
```

```{r plot}
require(reshape)
require(ggplot2)
t10 <- subset(top10,select=cols)
t10<-melt(t10,id=c('EVTYPE'),variable_name = 'HEALTHIMPACT')

p <- ggplot(t10, aes(x=as.factor(t10$EVTYPE), y= value, fill=t10$HEALTHIMPACT)) +
    geom_bar( stat = "identity") +
    coord_flip() 
p <- p + xlab("Event Type")
p <- p +  ylab("Incidents")
p <- p + ggtitle("USA Deadliest Severe Event Types 1996-2010")
p <- p + guides(fill=guide_legend(title=""))
p

```

## Economic Impacts

```{r economic}
EXP 
```

```{r old}

cols <- c('STATE', 'EVTYPE', 'BGN_DATE', 'END_DATE', 'COUNTYNAME','FATALITIES','INJURIES','PROPDMG', 'PROPDMGEXP','CROPDMG','CROPDMGEXP','REMARKS')
```